### 2022-08-21 background/background.tex
update the final version

\label{ch:background}
In this chapter, the background covered in this thesis is presented according to the following structure: in Section~\ref{ch3:dnn} and Section~\ref{ch3:bp}, we first briefly discuss deep neural network (DNN), back-propagation and optimisation methods. Then, in Section~\ref{ch3:rl}, some background to reinforcement learning (RL) is introduced. Next, in Section~\ref{ch3:policy-optim}, we discuss more details about policy optimisation in deep reinforcement learning (DRL) and related DRL algorithms that are covered in this thesis. Finally, in Section~\ref{ch3:dpp}, we introduce the fundamentals of determinantal point processes (DPPs).
%Finally, in Section~\ref{ch3:env}, we introduce the benchmark environments which are adopted to evaluate the proposed methods.

\section{Deep Neural Network in DRL}
\begin{figure}[h]
    \centering
    \includegraphics[width=0.7\textwidth]{figures/background/mlp_background.pdf}
    \caption{A simple structure of a multi-layer perceptron (MLP). It consists of an input layer, two hidden layers and an output layer. The highlighted lines (\textcolor{red}{red}) show the connection between a single neuron and its previous layer.}
    \label{fig:nn}
\end{figure}
\label{ch3:dnn}
\subsection{Multi-layer Perceptron}
Deep learning is based on research of artificial neural networks (ANNs). The structure of ANNs is inspired by animal brains, and mimics signal propagation between biological neurons. A feed-forward neural network (FFNN) or multi-layer perceptron (MLP) is a standard structure of ANN, and it consists of several neuron layers, which include an input layer, one or several hidden layers and an output layer (see Figure~\ref{fig:nn}). In each layer, there are multiple neurons, and each neuron is connected to those in the previous layer with corresponding weights $w_{i}$ and a bias $b$. For example, in Figure~\ref{fig:nn}, the transmission of the signal between the highlighted neuron and the previous layer can be expressed as:
\begin{equation}
    f = \sigma (w_{0}x_{0} + w_{1}x_{1} + w_{2}x_{2} + b)
\label{eq:nn_compute}
\end{equation}
here, $\sigma(\cdot)$ represents an activation function, such as a rectifier linear unit (ReLU)~\cite{nair2010rectified}, sigmoid$(\cdot)$ and tanh$(\cdot)$, which are used to add non-linearity to the network, because most real-world problems are nonlinear. The hidden layers in a neural network are used to learn various non-linear combinations of the input data, and the learned non-linear combinations are also called features. These features are the key factors to decide the prediction of the network.

The general forward propagation in an MLP that can be described as the following way. Firstly, the input layer receives the input data of a specific task. The input data can be a flattened image or the property of a house (e.g., the number of rooms, location, and area). Then, in each hidden layer, the output of each neuron is computed according to Equation~\eqref{eq:nn_compute}, and this computation proceeds from the first hidden layer to the last hidden layer. Finally, in the output layer, the output of the network can be the probability of each class in a classification task, the predicted value in a regression task, or the policy/predicted state value in a DRL task.

However, the initial parameters (i.e., weights and bias) of a FFNN may not fit a function for solving a specific task. Thus, the parameters need to be updated through a back-propagation method which is discussed in Section~\ref{ch3:bp}.
% CNN
\subsection{Convolutional Neural Network}
Multi-layer perceptrons (MLPs) are somewhat inefficient in dealing with computer vision tasks because of its fully connected structure. As the input size increases, the number of parameters (weight and bias) also increases. Convolutional neural network (CNN)~\cite{lecun1998gradient} alleviates this problem by using learnable kernels/filters, where each filter traverses the entire image. \td{CNNs are a strong inductive bias that reflects translation invariance of the physical world.} Compared with an MLP, the parameters of these kernels are shared, less, and easy to train. In previous works~\cite{krizhevsky2012imagenet,girshick2014rich,detone2016deep}, CNN has demonstrated its better feature representation ability than other hand-crafted methods, such as histogram of oriented gradient (HOG)~\cite{dalal2005histograms} and scale-invariant feature transform (SIFT)~\cite{lowe2004distinctive}. Nowadays, an increased number of powerful network structures have been proposed, such as the VGG network~\cite{simonyan2014very}, the residual network~\cite{he2016deep}, the dense network~\cite{huang2017densely}, and the capsule network~\cite{sabour2017dynamic}. These network structures improve state-of-the-art performance on various benchmark tasks~\cite{deng2009imagenet,everingham2010pascal}. In DRL, CNNs are regularly used to extract features from high-dimensional inputs (e.g., images) for further processing. In this thesis, as in many DRL studies, we put less effort into the design of the network structure.
%\td{[shifting and scaling in-variance makes CNN important]}

\section{Back-propagation and Optimisation}
\label{ch3:bp}
Gradient descent~\cite{bottou2010large} is a common technique used to update the parameters of the deep neural network (DNN) to fit a model, and it relies on the gradient of an objective function/loss function with respect to the parameters of the network. The gradient can be calculated efficiently through using the back-propagation algorithm~\cite{kelley1960gradient}. In the forward propagation of an MLP with $N$ layers, the output of each hidden layer can be formulated as:
\begin{equation}
    \mathbf{f}_{i} = \sigma(\mathbf{W}_{i-1}\mathbf{f}_{i-1} + \mathbf{b}_{i-1}), \quad i = 1, 2, \cdots, N,
\end{equation}
where $\mathbf{f}_{i}$ is the output of the current hidden layer, $\mathbf{f}_{i-1}$ is the output of the previous hidden layer, and the input is $\mathbf{f}_{0}$; $\mathbf{W}_{i-1}$ and $\mathbf{b}_{i-1}$ are the weights and bias of the current layer, respectively. Then, in this example, we use the squared error as the loss function to calculate the gradient and update the parameters of the network:
\begin{equation}
    \min_{\bm{\theta}}\mathcal{L} = \norm{\mathbf{y} - \mathbf{f}_{N}(\mathbf{x};\bm{\theta})}^{2},
\end{equation}
where $\bm{\theta}$ is the collection of parameters of the network: $\{\mathbf{W_{0}}, \mathbf{b_{0}}, \mathbf{W_{1}}, \mathbf{b_{1}}, \cdots, \mathbf{W_{N-1}}, \mathbf{b_{N-1}}\}$, $\mathbf{y}$ is the ground truth value and $\mathbf{x}$ is the input data. To minimise the error between the ground truth and the prediction, we need to compute the partial derivatives of $\mathcal{L}$ with respect to the parameters $\bm{\theta}_{i}$ of each layer using the chain rule:
\begin{equation}
   \nabla_{\bm{\theta}_{i}}\mathcal{L} = \frac{\partial \mathcal{L}}{\partial \bm{\theta_{i}}} = \frac{\partial \mathcal{L}}{\partial \mathbf{f}_{N}} \frac{\partial\mathbf{f}_{N}}{\partial \mathbf{f}_{N-1}}\cdots \frac{\partial\mathbf{f}_{i+2}}{\partial \mathbf{f}_{i+1}} \frac{\partial\mathbf{f}_{i+1}}{\partial \bm{\theta}_{i}}.
\end{equation}

Once get the gradient $\nabla_{\bm{\theta}}\mathcal{L}$, the parameters $\bm{\theta}$ can be updated in the opposite direction of the gradient with a step size $\eta$ (i.e. learning rate) via gradient descent:
\begin{equation}
    \bm{\theta} \leftarrow \bm{\theta} - \eta\nabla_{\bm{\theta}}\mathcal{L}.
\end{equation}
However, gradient descent requires the calculation of the gradient over all samples, and it is inefficient when the number of samples is large. Stochastic gradient descent (SGD) reduces computational complexity by randomly splitting the entire samples into several minibatches, and each minibatch is comprised of one or more samples. Then, the gradient is computed over each minibatch samples to update the parameters. 
%Furthermore, SGD with momentum~\cite{qian1999momentum} can accelerate the convergence speed on top of standard SGD.
In this thesis, two variants of the SGD method are used to update the parameters of the network: root mean square propagation (RMSProp)~\cite{tieleman2012lecture} and Adam~\cite{kingma2014adam}.

Root mean square propagation (RMSProp) can adaptively adjust the learning rate for different parameters (i.e., per-parameter learning rate) to accelerate the convergence, the update rule can be written as:
\begin{align}
    \mathbf{v} &\leftarrow \beta \mathbf{v} + (1 - \beta)\nabla_{\bm{\theta}}\mathcal{L}\odot\nabla_{\bm{\theta}}\mathcal{L} \label{eq:squared_grad}\\
     \bm{\theta} &\leftarrow \bm{\theta} - \frac{\eta}{\epsilon+\sqrt{\mathbf{v}}}\odot\nabla_{\bm{\theta}}\mathcal{L} \label{eq:rmsprop_update}.
\end{align}
In Equation~\eqref{eq:squared_grad}, $\mathbf{v}$ is the moving average of the squared gradient, $\odot$ is the element-wise multiplication, and $\beta$ is the decay coefficient which indicates the ratio of history information used to compute the average value. In Equation~\eqref{eq:rmsprop_update}, $\epsilon$ is a small value to prevent division by zero, $\frac{1}{(\epsilon + \sqrt{\mathbf{v}})}$ is the weight to balance the learning rate for individual parameter, in doing so, the learning rate for large gradient will decrease to avoid exploding, and the learning rate for small gradient will increase to avoid vanishing.

Adam is a combination of momentum~\cite{qian1999momentum} and RMSProp, and it is the most popular optimiser in DRL community. The update rule of Adam is:
\begin{align}
    \mathbf{m} &\leftarrow \beta_{1} \mathbf{m} + (1 - \beta_{1})\nabla_{\bm{\theta}}\mathcal{L} \label{eq:momentum_adam}\\
    \mathbf{v} &\leftarrow \beta_{2} \mathbf{v} + (1 - \beta_{2})\nabla_{\bm{\theta}}\mathcal{L}\odot\nabla_{\bm{\theta}}\mathcal{L} \label{eq:squared_grad_adam}\\
     \bm{\theta} &\leftarrow \bm{\theta} - \frac{\eta}{\epsilon+\sqrt{\mathbf{v}}}\mathbf{m} \label{eq:adam_update}.
\end{align}
Equation~\eqref{eq:momentum_adam} is the momentum term, $\mathbf{m}$ is the running average of the gradient, and $\beta_{1}$ is the decay coefficient. The momentum term can correct the gradient in the right direction and thus accelerate the convergence. The other parts are similar to RMSProp.

\section{Reinforcement Learning}
\label{ch3:rl}
\subsection{Markov Decision Process}
\begin{figure}[h]
    \centering
    \includegraphics[width=0.8\textwidth]{figures/background/mdp_background.pdf}
    \caption{An overview of reinforcement learning (RL).}
    \label{fig:rl_desc}
\end{figure}
Reinforcement learning (RL) can be formulated under the framework of a Markov Decision Process (MDP); it is used to learn an optimal policy to solve sequential decision-making problems. The MDP is a mathematical framework which describes a fully observable environment in RL, and it is composed of a 5-tuple $<\mathcal{S}, \mathcal{A}, \mathcal{P}, \mathcal{R}, \gamma>$. $\mathcal{S}$ is a set of states, $\mathcal{A}$ is a set of actions, $\mathcal{P}$ is a transition probability/dynamics function: $p(s_{t+1}|s_{t},a_{t})$ which only depends on the current state and action, $\mathcal{R}$ is a reward function, and $\gamma$ is a discount factor. At each timestep $t$, a state $s_{t}\in\mathcal{S}$ is received by an agent from the environment (see Figure~\ref{fig:rl_desc}). An action $a_{t}\in\mathcal{A}$ is sampled by the agent according to its policy $\pi(a_{t}|s_{t})$. Then, the next state $s_{t+1}$ and reward $r_{t}$ are provided to the agent according to the transition function and the reward function. The goal of the RL is to have the agent learn a policy that maximises the expected return $\mathbb{E}_{\pi}[G_{t}]$~\cite{sutton2018reinforcement}:
\begin{equation}
\centering
    G_{t} =\sum_{k=0}^{T-1}\gamma^{k}r_{t+k}.
\label{eq:acc_reward_func}
\end{equation}
Here, the discount factor $\gamma$ exponentially downplays the influence of future rewards. {In a robot manipulation task, the state $s_{t}$ can be the velocity and position of each joint of a robot arm. The action $a_{t}$ can be the velocities of actuators (control signals)} and the reward $r_{t}$ might be calculated based on the distance between the gripper of the robot arm and the target position.

\subsection{Goal-Oriented Reinforcement Learning}
\label{ch3:goal_rl}
The purpose of goal-oriented RL or multi-goal RL is to acquire a policy that can achieve different goals within the environment. 
%Universal Value Function Approximators (UVFA) is proposed to solve such 
In this thesis, the experiments in Chapter~\ref{ch:esil} and Chapter~\ref{ch:dtgsh} follow the terminology suggested by OpenAI~\cite{plappert2018multi}, in which the possible goals are drawn from a goal space $\mathcal{G}$, and the goal being pursued does not influence the environment dynamics. Two types of goal are recognised. One is the desired goal $g\in \mathcal{G}$, which is the target position {or state}, and may be different for different episodes. Within a single episode, $g$ is constant.  The second type of goal is the achieved goal $g^{ac}_{t}$, which is the achieved state in the environment (e.g., the position of an object), and this is considered to be different at each timestep in an episode. In an episode, each transition can be represented as $(s_t|\langle g, g^{ac}_{t} \rangle, a_t, r_t, s_{t+1}|\langle g, g^{ac}_{t+1} \rangle)$, where $s_t$ indicates a state, $a_t$ indicates an action and $r_t$ indicates a~reward; $\langle \, , \rangle$ is simply used to represent the grouping of goals.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.85\textwidth]{figures/background/her_background.pdf}
    \caption{A brief illustration of the hindsight goal relabelling in a robotic manipulation task~\cite{plappert2018multi}. In this task, the agent needs to control the gripper to hit the puck and make it slide to a specific position, where the achieved goal is the position of the puck and the desired goal is the target position (the red spot).}
    \label{fig:her_back}
\end{figure}
In sparse reward settings, an agent will only get non-negative rewards when the desired goal $g$ is achieved. The sparse reward function can be defined as:
\begin{equation}
r_{t}\left(g^{ac}_{t+1}, g\right):=
\begin{cases}
0& \text{if $\left\|g^{ac}_{t+1} - g\right\|\leq\epsilon$}\\
-1& \text{otherwise},
\end{cases}
\label{eq:sparse_reward}
\end{equation}
where $\epsilon$ is a threshold value, used to identify if the agent has achieved the goal. {However, the desired goal, $g$, might be difficult to reach during training. Thus, hindsight experiences~\cite{andrychowicz2017hindsight} (see Figure~\ref{fig:her_back}) are created through replacing the original desired goal $g$ with the current achieved goal $g_{t+1}^{ac}$ to augment successful samples, and then reward $r_{t}$ can be recomputed according to Equation~\eqref{eq:sparse_reward}. The modification of the desired goal can be denoted as $g^\prime$ and transitions from hindsight experiences can be represented as $(s_t|\langle g^\prime, g^{ac}_{t} \rangle, a_t, r_t, s_{t+1}|\langle g^\prime, g^{ac}_{t+1} \rangle)$.} {Intuitively, introducing $g^\prime$ serves a useful purpose in the early stages of training; taking, for example, a robot reaching task, the agent has no prior ``concept" of how to move its effector to a specific location in space. Thus, even these original failed episodes contain valuable information for ultimately learning a useful control policy for the original, desired goal $g$.}

@@

\subsection{Value-based Methods}
\label{ch3:value-based}
% value based
Value-based methods aim to learn a state or state--action value function instead of an explicit policy. The value function can evaluate the quality of a given state or the quality of an action performed at a given state. Then, the agent can select an action according to the corresponding estimate. This section will discuss how to derive a policy through the value function.

>>>>

\label{ch:background}
In this chapter, the background covered in this thesis is presented according to the following structure: in Section~\ref{ch3:dnn} and Section~\ref{ch3:bp}, we first briefly discuss deep neural networks (DNNs), back-propagation and optimisation methods. Then, in Section~\ref{ch3:rl}, some background to reinforcement learning (RL) is introduced. Next, in Section~\ref{ch3:policy-optim}, we discuss more details about policy optimisation in deep reinforcement learning (DRL) and related DRL algorithms that are covered in this thesis. Finally, in Section~\ref{ch3:dpp}, we introduce the fundamentals of determinantal point processes (DPPs).
%Finally, in Section~\ref{ch3:env}, we introduce the benchmark environments which are adopted to evaluate the proposed methods.

\section{Deep Neural Network in DRL}
\begin{figure}[h]
    \centering
    \includegraphics[width=0.9\textwidth]{figures/background/NN.pdf}
    \caption[A simple Structure of an MLP.]{A simple structure of a multi-layer perceptron (MLP). It consists of an input layer, two hidden layers and an output layer. The highlighted lines (red) show the connection between a single neuron and its previous layer.}
    \label{fig:nn}
\end{figure}
\label{ch3:dnn}
\subsection{Multi-layer Perceptron}
Deep learning is based on research of artificial neural networks (ANNs). The structure of ANNs is inspired by animal brains, and mimics signal propagation between biological neurons. A feed-forward neural network (FFNN) or multi-layer perceptron (MLP) is a standard structure of ANN, and it consists of several neuron layers, which include an input layer, one or several hidden layers and an output layer (see Figure~\ref{fig:nn}). In each layer, there are multiple neurons. Consider a single neuron connected to those in the previous layer with weights $w_{i}$ and a bias $b$. For example, in Figure~\ref{fig:nn}, the transmission of the signal between the highlighted neuron and the previous layer can be expressed in a general form as:
% \begin{equation}
%     f = \sigma (w_{0}x_{0} + w_{1}x_{1} + w_{2}x_{2} + b)
% \label{eq:nn_compute}
% \end{equation}
\begin{equation}
    %f = \sigma (w_{0}x_{0} + w_{1}x_{1} + w_{2}x_{2} + b)
    f = \sigma (b + \sum_{i=0}^{N} w_{i}x_{i})
\label{eq:nn_compute}
\end{equation}
where, $N$ is the number of input neurons from the previous layer and $N$ is 3 in the above example. $\sigma(\cdot)$ represents an activation function, such as a rectifier linear unit (ReLU)~\cite{nair2010rectified}, sigmoid$(\cdot)$ and tanh$(\cdot)$, which are used to add non-linearity to the network, because most real-world problems are nonlinear. The hidden layers in a neural network are used to learn various non-linear combinations of the input data, and the learned non-linear combinations are sometimes called features. These features are the key factors to decide the prediction of the network.

The general forward propagation in an MLP that can be described as the following way. Firstly, the input layer receives the input data of a specific task. The input data can be a flattened image or the property of a house (e.g., the number of rooms, location, and area). Then, in each hidden layer, the output of each neuron is computed according to Equation~\eqref{eq:nn_compute}, and this computation proceeds from the first hidden layer to the last hidden layer. Finally, in the output layer, the output of the network can be the preferences of each class in a classification task, the predicted value in a regression task, or the policy/predicted state value in a DRL task.

However, the initial parameters (i.e., weights and bias) of a FFNN may not fit a function for solving a specific task. Thus, the parameters need to be updated through a back-propagation method which is discussed in Section~\ref{ch3:bp}.
% CNN
%As the input size increases, the number of parameters (weight and bias) also increases.
\subsection{Convolutional Neural Network}
Multi-layer perceptrons (MLPs) are somewhat inefficient in dealing with computer vision tasks because of their fully connected structures. When the MLP is used to classify an image, each node in the hidden layer (the next layer of the input layer) is connected to all pixels of the input image. Thus, there will be a large number of parameters in the MLP when the size of the input image is large. For example, when the size of the input image is $256\times256$, the first hidden layer has $256\times256\times N$ parameters to learn, and $N$ is the number of neurons in this layer. Therefore, it is essential to find an approach to substantially reduce parameters. Convolutional neural networks (CNNs)~\cite{lecun1998gradient} alleviate this problem by employing the parameter sharing strategy and local connectivity. Instead of connecting each node to all nodes in the previous layer, CNNs use a set of learnable filters with relatively small size (e.g. $9\times9$) to slide across the entire image to produce the features. In this case, the first convolution layer only has $9\times9\times M$ parameters to learn, and $M$ is the number of filters, therefore CNNs greatly reduce the number of parameters. Benefiting from parameter sharing, CNNs might have an implicit strong inductive bias that reflects translation invariance in some desired mapping from physical world to feature space. Compared with an MLP, the parameters of these kernels are shared, fewer in number, and easier to train. In previous works~\cite{krizhevsky2012imagenet,girshick2014rich,detone2016deep}, CNN has demonstrated its better feature representation ability than other hand-crafted methods, such as histogram of oriented gradient (HOG)~\cite{dalal2005histograms} and scale-invariant feature transform (SIFT)~\cite{lowe2004distinctive}. Nowadays, an increased number of powerful network structures have been proposed, such as the VGG network~\cite{simonyan2014very}, the residual network~\cite{he2016deep}, the dense network~\cite{huang2017densely}, and the capsule network~\cite{sabour2017dynamic}. These network structures improve state-of-the-art performance on various benchmark tasks~\cite{deng2009imagenet,everingham2010pascal}. In DRL, CNNs are regularly used to extract features from high-dimensional inputs (e.g., images) for further processing. In this thesis, as in many DRL studies, we put less effort into the design of the network structure.
%\td{[shifting and scaling in-variance makes CNN important]} 
%\newpage
%otherwise it may cause aliasing of the input.

\section{Back-propagation and Optimisation}
\label{ch3:bp}
Gradient descent~\cite{bottou2010large} is a common technique used to update the parameters of the deep neural network (DNN) to fit a model, and it relies on the gradient of an objective function/loss function with respect to the parameters of the network. The gradient can be calculated efficiently through the back-propagation algorithm~\cite{kelley1960gradient}. In the forward propagation of an MLP with $N$ layers, the output of each hidden layer can be formulated as:
\begin{equation}
    \mathbf{f}_{i} = \sigma(\mathbf{W}_{i-1}\mathbf{f}_{i-1} + \mathbf{b}_{i-1}), \quad i = 1, 2, \cdots, N,
\end{equation}
where $\mathbf{f}_{i}$ is the output of the current hidden layer, $\mathbf{f}_{i-1}$ is the output of the previous hidden layer, and the input is $\mathbf{f}_{0}$; $\mathbf{W}_{i-1}$ and $\mathbf{b}_{i-1}$ are the weights and bias of the current layer, respectively. Then, in this example, we use the squared error as the loss function to calculate the gradient and update the parameters of the network:
\begin{equation}
    \min_{\bm{\theta}}\mathcal{L} = \norm{\mathbf{y} - \mathbf{f}_{N}(\mathbf{x};\bm{\theta})}^{2},
\end{equation}
where $\bm{\theta}$ is the collection of parameters of the network: $\{\mathbf{W_{0}}, \mathbf{b_{0}}, \mathbf{W_{1}}, \mathbf{b_{1}}, \cdots, \mathbf{W_{N-1}}, \mathbf{b_{N-1}}\}$, $\mathbf{y}$ is the ground truth value and $\mathbf{x}$ is the input data. To minimise the error between the ground truth and the prediction, we need to compute the partial derivatives of $\mathcal{L}$ with respect to the parameters $\bm{\theta}_{i}$ of each layer using the chain rule:
\begin{equation}
   \nabla_{\bm{\theta}_{i}}\mathcal{L} = \frac{\partial \mathcal{L}}{\partial \bm{\theta_{i}}} = \frac{\partial \mathcal{L}}{\partial \mathbf{f}_{N}} \frac{\partial\mathbf{f}_{N}}{\partial \mathbf{f}_{N-1}}\cdots \frac{\partial\mathbf{f}_{i+2}}{\partial \mathbf{f}_{i+1}} \frac{\partial\mathbf{f}_{i+1}}{\partial \bm{\theta}_{i}}.
\end{equation}

Once we get the gradient $\nabla_{\bm{\theta}}\mathcal{L}$, the parameters $\bm{\theta}$ can be updated in the opposite direction of the gradient with a step size $\eta$ (i.e. learning rate) via gradient descent:
\begin{equation}
    \bm{\theta} \leftarrow \bm{\theta} - \eta\nabla_{\bm{\theta}}\mathcal{L}.
\end{equation}
However, gradient descent requires the calculation of the gradient over all samples, and it is inefficient when the number of samples is large. Stochastic gradient descent (SGD) reduces computational complexity by randomly splitting the entire samples into several minibatches, and each minibatch is comprised of one or more samples. Then, the gradient is computed over each minibatch samples to update the parameters. 
%Furthermore, SGD with momentum~\cite{qian1999momentum} can accelerate the convergence speed on top of standard SGD.
In this thesis, two variants of the SGD method are used to update the parameters of the network: root mean square propagation (RMSProp)~\cite{tieleman2012lecture} and Adam~\cite{kingma2014adam}.

Root mean square propagation (RMSProp) can adaptively adjust the learning rate for different parameters (i.e., per-parameter learning rate) to accelerate the convergence, and the update rule can be written as:
\begin{align}
    \mathbf{v} &\leftarrow \beta \mathbf{v} + (1 - \beta)\nabla_{\bm{\theta}}\mathcal{L}\odot\nabla_{\bm{\theta}}\mathcal{L} \label{eq:squared_grad}\\
     \bm{\theta} &\leftarrow \bm{\theta} - \frac{\eta}{\epsilon+\sqrt{\mathbf{v}}}\odot\nabla_{\bm{\theta}}\mathcal{L} \label{eq:rmsprop_update}.
\end{align}
In Equation~\eqref{eq:squared_grad}, $\mathbf{v}$ is the moving average of the squared gradient, $\odot$ is element-wise multiplication, and $\beta$ is the decay coefficient which indicates the ratio of history information used to compute the average value. In Equation~\eqref{eq:rmsprop_update}, $\epsilon$ is a small value to prevent division by zero, $\frac{1}{(\epsilon + \sqrt{\mathbf{v}})}$ is the weight to balance the learning rate for individual parameter, in doing so, the learning rate for a large gradient will decrease to avoid exploding, and the learning rate for small gradient will increase to avoid vanishing.

Adam is a combination of momentum~\cite{qian1999momentum} and RMSProp, and is the most popular optimiser in the DRL community. The update rule of Adam is:
\begin{align}
    \mathbf{m} &\leftarrow \beta_{1} \mathbf{m} + (1 - \beta_{1})\nabla_{\bm{\theta}}\mathcal{L} \label{eq:momentum_adam}\\
    \mathbf{v} &\leftarrow \beta_{2} \mathbf{v} + (1 - \beta_{2})\nabla_{\bm{\theta}}\mathcal{L}\odot\nabla_{\bm{\theta}}\mathcal{L} \label{eq:squared_grad_adam}\\
     \bm{\theta} &\leftarrow \bm{\theta} - \frac{\eta}{\epsilon+\sqrt{\mathbf{v}}}\mathbf{m} \label{eq:adam_update}.
\end{align}
Equation~\eqref{eq:momentum_adam} is the momentum term, $\mathbf{m}$ is the running average of the gradient, and $\beta_{1}$ is the decay coefficient. The momentum term can correct the gradient in the right direction, and thus accelerate the convergence. The other parts are similar to RMSProp.

\section{Reinforcement Learning}
\label{ch3:rl}
\subsection{Markov Decision Process}
\begin{figure}[h]
    \centering
    \includegraphics[width=0.8\textwidth]{figures/background/mdp_background.pdf}
    \caption{An overview of reinforcement learning (RL).}
    \label{fig:rl_desc}
\end{figure}
Reinforcement learning (RL) can be formulated under the framework of a Markov Decision Process (MDP); it is used to learn an optimal policy to solve sequential decision-making problems. The MDP is a mathematical framework which describes a fully observable environment in RL, and it is composed of a 5-tuple $<\mathcal{S}, \mathcal{A}, \mathcal{P}, \mathcal{R}, \gamma>$. $\mathcal{S}$ is a set of states, $\mathcal{A}$ is a set of actions, $\mathcal{P}$ is a transition probability/dynamics function: $p(s_{t+1}|s_{t},a_{t})$ which only depends on the current state and action, $\mathcal{R}$ is a reward function, and $\gamma$ is a discount factor. At each timestep $t$, a state $s_{t}\in\mathcal{S}$ is received by an agent from the environment (see Figure~\ref{fig:rl_desc}). An action $a_{t}\in\mathcal{A}$ is sampled by the agent according to its policy $\pi(a_{t}|s_{t})$. Then, the next state $s_{t+1}$ and reward $r_{t}$ are provided to the agent according to the transition function and the reward function. The goal of the RL is to have the agent learn a policy that maximises the expected return $\mathbb{E}_{\pi}[G_{t}]$~\cite{sutton2018reinforcement}:
\begin{equation}
\centering
    G_{t} =\sum_{k=0}^{T-1}\gamma^{k}r_{t+k}.
\label{eq:acc_reward_func}
\end{equation}
Here, the discount factor $\gamma$ exponentially diminishes the influence of future rewards. {In a robot manipulation task, the state $s_{t}$ can be the velocity and position of each joint of a robot arm. The action $a_{t}$ can be the velocities of actuators (control signals)} and the reward $r_{t}$ might be calculated based on the distance between the gripper of the robot arm and the target position.

\subsection{Goal-Oriented Reinforcement Learning}
\label{ch3:goal_rl}
The purpose of goal-oriented RL or multi-goal RL is to acquire a policy that can achieve different goals within the environment. 
%Universal Value Function Approximators (UVFA) is proposed to solve such 
In this thesis, the experiments in Chapter~\ref{ch:esil} and Chapter~\ref{ch:dtgsh} follow the terminology suggested by OpenAI~\cite{plappert2018multi}, in which the possible goals are drawn from a goal space $\mathcal{G}$, and the goal being pursued does not influence the environment dynamics. Two types of goals are recognised. One is the desired goal $g\in \mathcal{G}$, which is the target position {or state}, and may be different for different episodes. Within a single episode, $g$ is constant.  The second type of goal is the achieved goal $g^{ac}_{t}$, which is the achieved state in the environment (e.g., the position of an object), and this is considered to be different at each timestep in an episode. In an episode, each transition can be represented as $(s_t|\langle g, g^{ac}_{t} \rangle, a_t, r_t, s_{t+1}|\langle g, g^{ac}_{t+1} \rangle)$, where $s_t$ indicates a state, $a_t$ indicates an action and $r_t$ indicates a~reward; $\langle \, , \rangle$ is simply used to represent the grouping of goals. To be clarified, the desired goal $g$ should be part of the state in RL. However, in order to highlight the specificity of different goal positions in goal-oriented RL, we split the desired goal out of the state. During training or evaluation, the desired goal and state are still concatenated together as the input of the neural network.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.85\textwidth]{figures/background/her_background_modified.pdf}
    \caption[A brief illustration of Hindsight Experience Replay (HER).]{A brief illustration of the hindsight goal relabelling in a robotic manipulation task~\cite{plappert2018multi}. In this task, the agent needs to control the gripper to hit the puck and make it slide to a specific position, where the achieved goal is the position of the puck and the desired goal is the target position (the red spot).}
    \label{fig:her_back}
\end{figure}
In sparse reward settings, an agent will only get non-negative rewards when the desired goal $g$ is achieved. The sparse reward function can be defined as:
\begin{equation}
r_{t}\left(g^{ac}_{t+1}, g\right):=
\begin{cases}
0& \text{if $\left\|g^{ac}_{t+1} - g\right\|\leq\epsilon$}\\
-1& \text{otherwise},
\end{cases}
\label{eq:sparse_reward}
\end{equation}
where $\epsilon$ is a threshold value, used to identify if the agent has achieved the goal. {However, the desired goal, $g$, might be difficult to reach during training. Thus, hindsight experiences~\cite{andrychowicz2017hindsight} (see Figure~\ref{fig:her_back}) are created through replacing the original desired goal $g$ with the current achieved goal $g_{t+1}^{ac}$ to augment successful samples, and then reward $r_{t}$ can be recomputed according to Equation~\eqref{eq:sparse_reward}. The modification of the desired goal can be denoted as $g^\prime$ and transitions from hindsight experiences can be represented as $(s_t|\langle g^\prime, g^{ac}_{t} \rangle, a_t, r_t, s_{t+1}|\langle g^\prime, g^{ac}_{t+1} \rangle)$.}

Intuitively, introducing $g^\prime$ serves a useful purpose in the early stages of training; taking, for example, a robot reaching task, the agent has no prior ``concept" of how to move its effector to a specific location in space. Thus, even these original failed episodes contain valuable information for ultimately learning a useful control policy for the original, desired goal $g$.
\newpage

@@

\subsection{Value-based Methods}
\label{ch3:value-based}
% value based
Value-based methods aim to learn a state or state--action value function instead of an explicit policy. The value function can evaluate the quality of a given state (state value function) or the quality of an action performed at a given state (Q-value function). Then, the agent can select an action according to the corresponding value. This section will discuss how to derive a policy through the value function.
